\documentclass[letterpaper]{article}
\usepackage{fullpage}
\usepackage{epsfig}

\title{Socorro: modular programming in FORTRAN}
\author{Ross Lippert, et al}

\begin{document}

\maketitle

Current version of this document will be essays or rants
which may then be remixed into a coherent article.


\section{C++ vs. FORTRAN: opinions}

\begin{verbatim}
(there should be some section like this in the paper, I think
though just what form it should take I do not know)
\end{verbatim}

Engaging in a large-scale software project begs the question
``why not use C++?''  C++ has emerged as a standardized
programming language to support a variety of software abstraction
techniques without loss of efficiency.

One answer is inertia.  This project began as an attempt to refactor
an legacy plane-wave code into a more maintainable and extensible
piece of software.  Since the bulk of that code was written in
FORTRAN, is was easier just to keep the FORTRAN for next incarnation.

Another answer is audience.  The electronic structures community
remains wedded to FORTRAN.  It may not be incorrect to say FORTRAN on
SGI's.  The primary goal of Socorro is to be extensible and
understandable by outsiders, and if those outsiders are FORTRAN users,
then Socorro will be FORTRAN for now (though one might argue its the
grad students who do the programming anyway and they all know C++).

Although FORTRAN does not supply the support for all modern software
engineering practices, significant improvements in data abstraction
have been provided which, when augmented by some minor user
conventions, we exploit in Socorro to give nearly first class support
for user defined types.

\include{modularity}

\include{memory}

\include{socorro}


\end{document}

















